### 2016-08-23 analysis-note-v2/RTPC.tex
more updates to the second chapter

We measure the $TDC_{max/2}$ along the length of the RTPC to take into account the variations in the electric and magnetic fields. The results can be seen in figure \ref{fig:RunNumber_61551_TDCmax_Zslice}, where a clear variation of the drift speed along the RTPC was observed. \\ 
\begin{figure}[tbp]
\centering
\vspace{-0.1in}
\includegraphics[scale=0.35]{fig_rtpc/TDC_profile.png}
\caption{Time profile of the collected hits for the RTPC good tracks in one experimental run, run 61511. The $TDC_{max/2}$ is the time at which the dropping edge passes half the maximum number of hits. }
\label{fig:TDC_profile}
\end{figure}

@@

\begin{figure}[tp]
\begin{minipage}[c]{.46\linewidth}
\hspace*{-0.3in}
\includegraphics[height=9.5cm]{fig_rtpc/RunNumber_61551_TDCmax_Zslice.png}
\caption{$TDC_{max/2}$ variation as a function of the longitudinal position along the RTPC in one experimental run, 61510.} 
\label{fig:RunNumber_61551_TDCmax_Zslice}
\end{minipage} \hfill
\begin{minipage}[c]{.46\linewidth}
\hspace*{-0.3in}
\includegraphics[height=9.0cm]{fig_rtpc/TDCmax_ratio_final.png}
\caption{Ratio between $TDC_{max/2}$ extracted using the RTPC good tracks and the one extracted using the clean elastic events, both from the 1.204-GeV dataset.}
\label{fig:TDCmax_ratio}
\end{minipage}
\end{figure}

@@

As $TDC_{max/2} -15 $ ($= 
\frac{Drift~path~length}{Drift~speed}$) varies with both run number (time) and 
the geometry of the RTPC ($z$ along the RTPC). We
parametrize the dependence on $z$ for each run in the form:\\
\begin{equation}
TDC_{max/2} (z)= p_{0} +  p_{1} * e^{p_{2}*(z-p_{3})^{2}}.
\label{equ:drift_speed_fit}
\end{equation}
Then fits are performed for these parameters as functions of the run number, such that the drift speed parametrization depends on the run and $z$ along the RTPC. The numeric parametrizations of $p_{0}$, $p_{1}$, $p_{2}$, and $p_{3}$ can be found in Appendix \ref{app:RTPC_appendix}, table \ref{Table:TDCmax}. \\

@@

\begin{figure}[tp]
\hspace{-0.3in}
\includegraphics[scale=0.45]{fig_rtpc/Drift_run_number.png}
\caption{ The integrated $TDC_{max/2}$ over the full z-RTPC as a function of 
run number.  The gain after calibration in the collected good tracks (GT) and 
elastic events (El) are shown for a sample of runs.}
\label{fig:Drift_run_number}
\end{figure}

@@

The procedure steps are:\\
\begin{itemize}
\item We first select a sample of well identified elastic events in our experimental data. Using the
   kinematics of the scattered electrons in these events, we generate in GEANT4
   a sample of events identical to the measured ones. 
\item  For each couple of events (the experimental one and its simulated equivalent)
   we make a linear correlation between $R$, from simulation, and TDC from 
   real data, to link the GEANT4 hits to the measured ones. In our TPC, the 
   maximum $R$ equals 60 mm (close to the anode) at the trigger time, 
   (TDC$_{min}$ = 15 TDCs), and the minimum R equals 30 mm (close to the 
   cathode) at maximum TDC ($TDC_{max}$). Thus, the linear correlations can be 
   written as:
\begin{equation}
   R(TDC) = \frac{60-30}{15-TDC_{max}} (TDC-15) + 60.
\label{equ:R_TDC}
\end{equation}
We apply 3 TDCs-wide windows around the linear R-TDC correlation to associate the 
GEANT4 simulated hits to measured hits.

@@

\begin{figure}[tp]
\centering
\includegraphics[scale=0.45]{fig_rtpc/TdcPhi_p1_10.png}
\caption{(First pass). Distribution of $\Delta \phi$ as a function of TDC for 
   in one $z$ bin. The width of the bin is 10 mm, 
with the center at $z$ = 5 mm. The black line represents a fit for 
$\Delta\phi$.  }
\label{fig:1st_pass_delta_phi}
\end{figure}

@@

~~~~The second method requires improvements in the simulation, in order to match the real experiment. In summary, these improvements are:
\begin{itemize}
\item Implementation in the simulation of the previously extracted parametrizations of the drift speed and the drift paths.
\item Global ADC normalization to give reasonable simulated values. In the left module, 1 ADC is equal to 
17 eV, while it is equal to 21 eV in the right module. These normalizations were extracted from the comparison of the data to the GEANT4 simulation.
\item Rejection of the signals from bad pads. During the experiment, the readout system of the RTPC suffered from 555 dead or noisy pads. These pads are marked by the dotted squares in figure \ref{fig:pad_row_col}.
\item Smearing the position of the simulated hits. Experimentally, the average number of hits per track is around 80 while the initially reconstructed mean value from the simulation is around 50. We apply a Gaussian smearing on the position of the simulated hits to make the simulation more realistic, see figure \ref{fig:NNb_hits}.

\item Application of the TPC's DAQ cut on the simulated data. Experimentally, each pad must have at least 3 consecutive time bins, with ADC values above the threshold, 35 ADCs, in order to be recorded. Then, the hits of three neighboring bins on each side of the above threshold bins are also recorded, while the other hits are not.
\end{itemize}

\begin{figure}[tp]
\centering
\includegraphics[scale=0.40]{fig_rtpc/NNb_hits.png}
\caption{The distributions of the number of hits per track. The black, the blue and the red distributions are the number of hits per track  for respectively, the experimental data, simulation before smearing and the simulation with smearing.}
\label{fig:NNb_hits}
\end{figure}

To illustrate the procedure of comparing simulated tracks with real tracks,
we show in Figure \ref{fig:EVENT_adc_tdc} the simulated ADCs and the 
experimental ones as a function of the TDCs for the same track. 
Then, the gain of each pad can be obtained as the ratio of the mean experimental 
ADCs to the one of the simulated track. For each pad, these gain 
ratios were collected from all the identified elastic events and fitted by a 
Landau function to obtain their gains.

As mentionned earlier, we refine this calibration after 
the extracted gains are applied to the experimental data. We compare the ratio between the 
mean ADCs of each pad to the mean ADCs of the whole track. This ratio is 
collected from all the elastic events and a gain correction factor is extracted 
for each pad. The final gains of the second method are obtained with the application of 
this extra correction.\\

\begin{figure}[tp]
\centering
\vspace{-0.1in}
\includegraphics[scale=0.350]{fig_rtpc/EVENT_adc_tdc.png}
\caption{The simulated (top) and the experimental (bottom) ADCs versus TDC distributions for the same track. The same colors indicate hits that were registered in the same channel for the simulation and the experiment.}
\label{fig:EVENT_adc_tdc}
\end{figure}

As the two modules of the RTPC are electronically separated, we look at their calibrations separately. Figure \ref{fig:dedx_p_exp_2nd} shows the ratio between the calculated $\small{\frac{dE}{dX}}$, using the gains of both methods, and the GEANT4 simulated $\small{\frac{dE}{dX}}$ for the elastic events in the two modules of the RTPC. We conclude that extracting the gains from comparing data to simulation in terms of the ADCs of the individual tracks gives more precise gains than solving the series of equations. Figure \ref{fig:pad_row_col} shows the gains from the second method, where the dotted squares refer to the dead or noisy pads. These are the ones that are implemented in the EG6 reconstruction codes.\\

\begin{figure}[tbp]
\centering
\includegraphics[scale=0.400]{fig_rtpc/dedx_p_sim_exp_2nd.png}
\caption{The ratio between the experimental $dE/dx$, using the gains obtained with the two methods (1$^{st}$ in blue, and 2$^{nd}$ in black), to the GEANT4 simulated $dE/dx$, plotted for the two modules of the RTPC, respectively left and right. }
\label{fig:dedx_p_exp_2nd}
\end{figure}

@@

To check the validity of the extracted gains, we show $\small{\frac{dE}{dX}}$ versus the momentum (per charge unit) measured in the RTPC for all the collected tracks of the 1.206 GeV data set. In the plots, we add the theoretical lines derived from the Bethe-Bloch formula for possible detected particles: $^4He$, $^3He$, $^3H$, Deuterium (d) and the protons.
\begin{figure}[tbp]
\includegraphics[scale=0.35]{fig_rtpc/dedx_p_l.png}
\includegraphics[scale=0.35]{fig_rtpc/dedx_p_r.png}
\caption{The experimental $\small{\frac{dE}{dX}}$ calculated using the second method gains versus $p/q$ of all the collected good tracks in the RTPC in the early 1.204 GeV dataset.}
\label{fig:dedx_all}
\end{figure}
One can see the different bands corresponding to the different detected particles.  
Even though the bands are very wide, the $\small{\frac{dE}{dX}}$ can be used to 
perform particles identification for large data set and different physics 
processes. In this analysis, the $\small{\frac{dE}{dX}}$ is not used to 
identified the recoil $^{4}He$ in the coherent DVCS event selection, as the 
 set of exclusivity cuts that will be 
presented in chapter 4 appears to be strict enough.

>>>>

We measure the $TDC_{max/2}$ along the length of the RTPC to take into account 
the variations in the electric and magnetic fields. The results can be seen in 
figure \ref{fig:RunNumber_61551_TDCmax_Zslice}, where a clear variation of the 
drift speed along the RTPC was observed. \\ \begin{figure}[tbp]
\centering
\vspace{-0.1in}
\includegraphics[scale=0.35]{fig_rtpc/TDC_profile.png}
\caption{Time profile of the collected hits for the RTPC good tracks in one 
   experimental run, run 61511, integrated over the full longitudinal length 
(z) of the RTPC. The $TDC_{max/2}$ is the time at which the dropping edge 
passes half the maximum number of hits.}
\label{fig:TDC_profile}
\end{figure}

@@

\begin{figure}[tpb]
\begin{minipage}[c]{.46\linewidth}
\hspace*{-0.3in}
\includegraphics[height=9.5cm]{fig_rtpc/RunNumber_61551_TDCmax_Zslice.png}
 \caption{$TDC_{max/2}$ variation as a function of the longitudinal position 
 along the RTPC, $z$ position of the hit, in one experimental run, 61510.} 
 \label{fig:RunNumber_61551_TDCmax_Zslice}
\end{minipage} \hfill
\begin{minipage}[c]{.46\linewidth}
\hspace*{-0.3in}
\includegraphics[height=9.0cm]{fig_rtpc/TDCmax_ratio_final.png}
\caption{Ratio between $TDC_{max/2}$ extracted using the RTPC good tracks and 
the one extracted using the clean elastic events, both from the 1.204-GeV 
dataset, as function of the hit position ($z$).}
\label{fig:TDCmax_ratio}
\end{minipage}
\end{figure}

@@

As $TDC_{max/2} -15 $ ($= \frac{Drift~path~length}{Drift~speed}$) varies with 
both run number (time) and the geometry of the RTPC ($z$ along the RTPC). We
parametrize the dependence on $z$, hit (pad) position, for each run in the 
form:\\
\begin{equation}
TDC_{max/2} (z)= p_{0} +  p_{1} * e^{p_{2}*(z-p_{3})^{2}}.
\label{equ:drift_speed_fit}
\end{equation}
Then fits are performed for these parameters as functions of the run number, 
such that the drift speed parametrization depends on the run and $z$ along the 
RTPC. The numeric parametrizations of $p_{0}$, $p_{1}$, $p_{2}$, and $p_{3}$ 
can be found in Appendix \ref{app:RTPC_appendix}, table \ref{Table:TDCmax}. \\

@@

\begin{figure}[tpb]
\hspace{-0.3in}
\includegraphics[scale=0.9]{fig_rtpc/updates/TDCmax2-run-number.pdf}
\caption{ The integrated $TDC_{max/2}$ over the full z-RTPC as a function of 
run number.  The gain after calibration in the collected good tracks (GT) and 
elastic events (El) are shown for a sample of runs.}
\label{fig:Drift_run_number}
\end{figure}

@@

The procedure steps are:\\
\begin{itemize}
\item We first select a sample of well identified elastic events in our experimental data. Using the
   kinematics of the scattered electrons in these events, we generate in GEANT4
   a sample of events identical to the measured ones. 
\item  For each couple of events (the experimental one and its simulated equivalent)
   we make a linear correlation between $R$, from simulation, and TDC from 
   real data, to link the GEANT4 hits to the measured ones. In our TPC, the 
   maximum $R$ equals 60 mm (close to the anode) at the trigger time, 
   (TDC$_{min}$ = 15 TDCs), and the minimum R equals 30 mm (close to the 
   cathode) at maximum TDC ($TDC_{max}$). Thus, the linear correlations can be 
   written as:
\begin{equation}
   R(TDC) = \frac{60-30}{15-TDC_{max}} (TDC-15) + 60.
\label{equ:R_TDC}
\end{equation}
We apply 3 TDCs-wide windows around the linear R-TDC correlation to associate 
the GEANT4 simulated hits to measured hits. Figure \ref{fig:R_correlation} 
shows the distribution of $R$ versus TDC for the selected simulated hits in a 
single $z$ bin.

@@

\begin{figure}[tpb]
\centering
\includegraphics[scale=0.45]{fig_rtpc/updates/TdcR_check_p1_10.png}
\caption{(First pass). Distribution of $R$ as a function of TDC for the 
simulated hits located in one $z$ bin. The width of the bin is 10 mm, with the 
center at $z$ = 5 mm.}
\label{fig:R_correlation}
\end{figure}

\begin{figure}[tpb]
\centering
\includegraphics[scale=0.45]{fig_rtpc/TdcPhi_p1_10.png}
\caption{(First pass). Distribution of $\Delta \phi$ as a function of TDC for 
   in one $z$ bin. The width of the bin is 10 mm, 
with the center at $z$ = 5 mm. The black line represents a fit for 
$\Delta\phi$.  }
\label{fig:1st_pass_delta_phi}
\end{figure}

@@

~~~~The second method requires improvements in the simulation, in order to 
match the real experiment. In summary, these improvements are:
\begin{itemize}
\item Implementation in the simulation of the previously extracted 
   parametrizations of the drift speed and the drift paths.
\item Global ADC normalization to give reasonable simulated values. In the left 
   module, 1 ADC is equal to 17 eV, while it is equal to 21 eV in the right 
   module. These normalizations were extracted from the comparison of the data 
   to the GEANT4 simulation.
\item Rejection of the signals from bad pads. During the experiment, the 
   readout system of the RTPC suffered from 555 dead or noisy pads. These pads 
   are marked by the dotted squares in figure \ref{fig:pad_row_col}.
\item Smearing the position of the simulated hits. Experimentally, the average 
   number of hits per track is around 80 while the initially reconstructed mean 
   value from the simulation is around 50. We apply a Gaussian smearing on the 
   position of the simulated hits to make the simulation more realistic, see 
   figure \ref{fig:NNb_hits}.  
\item Application of the TPC's DAQ cut on the simulated data. Experimentally, 
   each pad must have at least 3 consecutive time bins, with ADC values above 
   the threshold, 35 ADCs, in order to be recorded. Then, the hits of three 
   neighboring bins on each side of the above threshold bins are also recorded, 
   while the other hits are not.
\end{itemize}

\begin{figure}[tpb]
\centering
\includegraphics[scale=0.35]{fig_rtpc/NNb_hits.png}
\caption{The distributions of the number of hits per track. The black, the blue 
and the red distributions are the number of hits per track  for respectively, 
the experimental data, simulation before smearing and the simulation with 
smearing.}
\label{fig:NNb_hits}
\end{figure}

To illustrate the procedure of comparing simulated tracks with real tracks,
we show in Figure \ref{fig:EVENT_adc_tdc} the simulated and the experimental 
ADCs as a function of the TDCs for the same track. Then, the gain of each pad 
is defined as the ratio between the mean recorded experimental ADCs to the mean 
simulated ADCs in the same track. Then for each pad, the gain ratios were 
collected from all the identified elastic tracks and these gains were fitted by 
a Landau function to give a first pass gain for each pad as the most probable 
value of the fit.

\begin{figure}[tpb]
\centering
\vspace{-0.1in}
\includegraphics[scale=0.320]{fig_rtpc/EVENT_adc_tdc.png}
\caption{The simulated (top) and the experimental (bottom) ADCs versus TDC 
distributions for the same track. The same colors indicate hits that were 
registered in the same channel for the simulation and the experiment.}
\label{fig:EVENT_adc_tdc}
\end{figure}

To make sure that the recorded ADCs by a given pad are similar to the recorded 
ADCs by other pads in the same track, we refined this calibration after the 
extracted gains were applied to the experimental data. At this step, we 
compared the mean ADCs of each pad to the mean ADCs of the whole track. This 
ratio is collected from all the elastic events and a gain correction factor is 
extracted and applied to get a final gain for each pad, labeled as second pass 
gain in figure \ref{fig:dedxre}.

As the two modules of the RTPC are electronically separated, we look at their 
calibrations separately. Figure \ref{fig:dedxre} shows the ratio between the 
calculated $\small{\frac{dE}{dX}}$, using the gains extracted from first and 
the second (both passes) meothods, and the GEANT4 simulated 
$\small{\frac{dE}{dX}}$ for the elastic events in the two modules of the RTPC.  
From the latter figure, one can see improvments obtained by using this method 
to fine tune the gains extracted only by comparing data to simulation. We 
conclude that extracting the gains from comparing data to simulation in terms 
of the ADCs of the individual tracks gives more precise gains than solving the 
series of equations. Figure \ref{fig:pad_row_col} shows the gains from the 
second method, where the dotted squares refer to the dead or noisy pads. These 
are the ones that are implemented in the EG6 reconstruction codes.\\

\begin{figure}[tpb]
\hspace{-0.5 cm}
\includegraphics[height=11.5cm]{fig_rtpc/updates/dedx_ratio_elastic_NAB.png}
\caption{The ratio between the experimental dEdx and the simulated ones for 
   different regions along $z$ for the two modules of the RTPC, right module on 
   top panel and left module on bottom panel. The blue lines are using the 
   first method of extracting the gains (comparing the experimental recorded 
   dEdx to the expected values calculated from the Bethe-Block formula). The 
green lines are using the fist pass of the second method (comparing simulation 
to data only) with all the elastic event sample. The brown lines are using the 
final gains extracted from the second method (both passes).}
\label{fig:dedxre}
\end{figure}

%\begin{figure}[tbp]
%\centering
%\includegraphics[scale=0.400]{fig_rtpc/dedx_p_sim_exp_2nd.png}
%\caption{The ratio between the experimental $dE/dx$, using the gains obtained 
%with the two methods (1$^{st}$ in blue, and 2$^{nd}$ in black), to the GEANT4 
%simulated $dE/dx$, plotted for the two modules of the RTPC, respectively left 
%nd right. }
%\label{fig:dedx_p_exp_2nd}
%\end{figure}

@@

To check the validity of the extracted gains, we show $\small{\frac{dE}{dX}}$ 
versus the momentum (per charge unit) measured in the RTPC for all the 
collected tracks of the 1.206 GeV data set. In the plots, we add the 
theoretical lines derived from the Bethe-Bloch formula for possible detected 
particles: $^4He$, $^3He$, $^3H$, Deuterium (d) and the protons.
\begin{figure}[tbp]
\includegraphics[scale=0.35]{fig_rtpc/updates/dedx_p_l.png}
\includegraphics[scale=0.35]{fig_rtpc/updates/dedx_p_r.png}
\caption{The experimental $\small{\frac{dE}{dX}}$ calculated using the second method gains versus $p/q$ of all the collected good tracks in the RTPC in the early 1.204 GeV dataset.}
\label{fig:dedx_all}
\end{figure}
One can see the different bands corresponding to the different detected particles.  
Even though the bands are very wide, the $\small{\frac{dE}{dX}}$ can be used to 
perform particles identification for large data set and different physics 
processes. In this analysis, the $\small{\frac{dE}{dX}}$ is not used to 
identified the recoil $^{4}He$ in the coherent DVCS event selection, as the 
 set of exclusivity cuts that will be 
presented in chapter 4 appears to be strict enough.
